**Measuring Depression and Anxiety with 4 items? Adaptation of the PHQ-4 to increase its Sensitivity to Subclinical Variability**

### 2025-06-05 paper/5_Collabra/manuscript.qmd
edits

This brief report aims at testing the possibility of enhancing - with minimal changes to the original scale - the PHQ-4 sensitivity to mild mood level inflections. In the first study, we will evaluate whether the new response option is prevalently used by participants, and whether it does capture a specific part of the latent measure. In the second study, we will compare the refined PHQ-4 version to the original one in terms of sensitivity to subclinical variability in depression, using the Beck Depression Inventory [BDI-II, @beck1996beck] and the State Trait Anxiety Inventory [STAI-5, @zsido2020development] as our ground-truth measures of depression and anxiety.

@@

Item Response Theory (IRT) provides insights into how well items and responses capture an underlying latent trait $\theta$. For each of the subscales, we fitted a unidimensional graded response model [GRM, @samejima1997graded]. For anxiety, the two items captured 89.2\% of the variance of the latent anxiety dimension ($\theta_{anxiety}$). The discrimination parameters suggested that the first item was less precise ($\alpha = 3.42$) than the second item ($\alpha = 12.55$) in its ability to discriminate between various levels of anxiety (i.e., each response on the second item covers a more exclusive range of $\theta_{anxiety}$, as can be seen in @fig-one). The two depression items captured 82.8\% of the variance of its latent trait ($\theta_{depression}$), and the opposite pattern was found: the first item had a higher precision ($\alpha = 16.46$) than the first ($\alpha = 2.41$). However, it is important to note that the "less precise" items were also the ones covering a larger portion of the latent space (being more sensitive especially on the lower end of the spectrum), offering an interesting trade-off between sensitivity and precision. Importantly for our objective, the added "Once or twice" option did cover a selective and unique portion of the latent space.

@@

One natural methodological limitation pertains to the interpretation of the latent dimensions in an IRT framework applied to pairs of items. In this study's context, references to for instance "the latent anxiety dimension" merely corresponds to the amalgamation of the two items of the anxiety subscale, and not to a more general and valid true anxiety factor.

>>>>

This brief report aims at testing the possibility of enhancing - with minimal changes to the original scale - the PHQ-4 sensitivity to mild mood level inflections. In the first study, we will evaluate whether the new response option is prevalently used by participants, and whether it does capture a specific part of the construct. In the second study, we will compare the refined PHQ-4 version to the original one in terms of sensitivity to subclinical variability in depression, using the Beck Depression Inventory [BDI-II, @beck1996beck] and the State Trait Anxiety Inventory [STAI-5, @zsido2020development] as our ground-truth measures of depression and anxiety.

@@

Item Response Theory (IRT) provides insights into how well items and responses capture an underlying latent trait $\theta$. For each of the subscales, we fitted a unidimensional graded response model [GRM, @samejima1997graded]. For anxiety, the two items captured 89.2\% of the variance of the latent anxiety dimension ($\theta_{anxiety}$). The discrimination parameters suggested that the first item was less precise ($\alpha = 3.42$) than the second item ($\alpha = 12.55$) in its ability to discriminate between various levels of anxiety (i.e., each response on the second item covers a more exclusive range of $\theta_{anxiety}$, as can be seen in @fig-one). The two depression items captured 82.8\% of the variance of its latent trait ($\theta_{depression}$), and the opposite pattern was found: the first item had a higher precision ($\alpha = 16.46$) than the second ($\alpha = 2.41$). However, it is important to note that the "less precise" items were also the ones covering a larger portion of the latent space (being more sensitive especially on the lower end of the spectrum), offering an interesting trade-off between sensitivity and precision. Most importantly for our objective, the added "Once or twice" option did cover a selective and unique portion of the latent space.

@@

**One natural methodological limitation pertains to the application of the IRT framework to pairs of items. While this is statistically sound (the graded model utilizes the full response pattern information and does not solely rely on the item covariance matrix for parameter estimation), it is important to underline that in our study's context, "the latent anxiety/depression dimension" merely corresponds to the amalgamation of the two items of the anxiety or depression subscale, and not to a more general and independent latent anxiety or depression factor.**

**In summary, the main take-away of this first study is that the "Once or twice" response option appears as a popular choice, which begs the question of its usefulness in capturing more fine-grained variations of the underlying dimensions as measured by independent tools.**
